---
title: "Reproducible Research Peer Assessment 1"
author: "JT"
date: "June 23, 2017"
output: 
  html_document:
    keep_md:
      true
---

## Loading and preprocessing the data

```{r read_data}
setwd("~/ds/ReproducibleResearch/Week2/RepData_PeerAssessment1")
data<-read.csv(unz("activity.zip","activity.csv"),stringsAsFactors = FALSE)
```
## What is mean total number of steps taken per day?

```{r exclude_na}
d_clean<-data[!is.na(data$steps),]
d_date<-aggregate(d_clean$steps,list(d_clean$date),mean)
names(d_date)<-c("date","steps")
hist(d_date$steps,main="Average Steps Per Day",xlab="Steps")
```

Summary measures for the average number of steps per day:
```{r summary_average_steps}
m<-mean(d_date$steps)
md<-median(d_date$steps)
```

The mean number of average steps per day is `r m`.  
The median number of average steps per day is `r md`.  

## What is the average daily activity pattern?

``` {r average_daily_pattern}
d_int<-aggregate(d_clean$steps,list(d_clean$interval),mean)
names(d_int)<-c("interval","steps")
with(d_int,plot(interval,steps,type="l",xlab="5-minute Interval",ylab="Steps",
                main="Average Number of Steps per Interval"))
```

So, which interval contains the maximum number of steps?
```{r calculate_max_steps}
max_index<-which.max(d_int$steps)
max_interval<-d_int[max_index,"interval"]
max_steps<-d_int[max_index,"steps"]
```

Interval `r max_interval` contains the maximum average number of steps: `r max_steps`.    

## Imputing missing values

```{r missing_data}
d_na<-data[is.na(data$steps),]
n_na<-dim(d_na)[1]
```

There are `r n_na` rows with NA in the data set.

Impute the missing step values using the average steps for the interval:

```{r impute_data}
# For each row d in d_na, set steps to average steps for the interval (stored in d_int)
d_na["steps"]<- d_int$steps[match(d_na$interval,d_int$interval)]
# Re-construct the data set by merging clean with imputed 
d_complete<-rbind(d_clean,d_na)
# Reorder
d_complete<-d_complete[order(as.numeric(row.names(d_complete))),]
# Confirm it all looks good
head(data,n=3)
d_int[d_int$interval %in% c(0,5,10),]
head(d_complete,n=3)
# Now check one data point for un-imputed data
data[800,]
d_complete[800,]
```

Now the summary measures and the  histogram
``` {r complete_hist}
d_complete_agg<-aggregate(d_complete$steps,list(d_complete$date),mean)
names(d_complete_agg)<-c("date","steps")
hist(d_complete_agg$steps,
     main="Average steps per day (Missing Data Imputed)",xlab="Steps")

```

``` {r summary_measures}
m<-mean(d_complete_agg$steps)
md<-median(d_complete_agg$steps)
```

The mean number of average steps per day is `r m`.  
The median number of average steps per day is `r md`. 
The means are identical, not so the median, which has changed a tiny bit.

As a double check
``` {r double_check}
matrix(c(sum(d_date$steps),sum(d_complete_agg$steps),dim(d_date)[1],dim(d_complete_agg)[1]),c(2,2))
all.equal(sum(d_complete_agg$steps)/dim(d_complete_agg)[1],
          sum(d_date$steps)/dim(d_date)[1])
```

## Are there differences in activity patterns between weekdays and weekends?

It appears so...
``` {r weekends}
library(lattice)
# Define a weekday vs. weekend function and a new column in the data
wd<-function(d){ d<-weekdays(as.Date(d))
    if (d=="Saturday"|d=="Sunday"){"Weekend"}else{"Weekday"}}
d_complete["weekday"]<-sapply(d_complete$date,wd)

# Aggregate by weekday/weekend and interval.  Plot.
d_final<-aggregate(d_complete$steps,list(d_complete$weekday,d_complete$interval),mean)
names(d_final)<-c("weekday","interval","steps")
xyplot(steps~interval|weekday,type="l",layout=c(1,2),data=d_final)

```
